The effect of prenatal environment on brain metabolic function and perception in a lizard

### 2025-09-07 ms.qmd
updates

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. However, early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects. Here, we investigated the combined influence of prenatal corticosterone (CORT) — a key stress hormone in reptiles — and incubation temperature on mitochondrial function and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta — two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta but not the olfactory bulbs. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms linking early-life stress, cell metabolism, and cognitive performance.

@@

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental inputs during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Long-lasting effects on energy production and oxidative stress in the brain may be a key mechanism by which cognitive function is impacted by developmental environments [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks correlate negatively with the number of round-shaped mitochondria — associated with higher oxidative stress — per presynaptic terminal. In conclusion, findings from different studies suggest that mitocondrial activity can be essential in shaping cognitive abilities. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

@@

## Discussion
We examined the impact of prenatal temperature and CORT on visual and chemical perception and determined whether metabolic function was the underlying reason for any perceptual differences we observed as has been shown in mammals [@hara_presynaptic_2014; @chaudhari2022early]. Our findings suggest that metabolic processes remained largely resilient to early environmental conditions. However, early-life conditions have region- and context-dependent effects on oxidative damage and prey detection. We observed interactive effects of prenatal CORT and incubation temperature on oxidative damage in the Optic Tecta, but no effects on perception of visual stimuli. In contrast, CORT decreased the time to perceive chemical stimuli despite the resilience of the Olfactory bulbs to prenatal conditions. Below, we discuss these findings in relation to mitochondrial metabolic function, oxidative stress, and behavioural performance.

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. In the leopard gecko (_Eublepharis macularius_), incubation temperature is an important determinant of metabolic capacity in some, but not all, brain nuclei [@coomber_independent_1997], while prenatal stress decreases mitochondrial respiration efficiency in the cortex, hippocampus, and hypotalamus of rats (_Rattus norvegicus_) [@gong_chronic_2011]. In contrast, metabolic processes in the optic tecta and olfactory bulbs of _L. delicata_ appear to be robust to early-life conditions, potentially to ensure that these brain regions can perform their essential functions. However, since the effects of early environment can be region-dependent [@coomber_independent_1997], it is possible that other brain regions not examined here may be more sensitive to prenatal CORT and temperature.

Alternatively, the effects of early environmental conditions on metabolic function and ROS production may be immediate rather than long-lasting. Mitochondria are known to be highly responsive to environmental changes and energetic demands [@gyllenhammer2020developmental]. It is possible that the metabolic effect of the early-life conditions dissipated once all animals were exposed to the same environment after hatching. Nonetheless, alterations in metabolic function during earlier stages would still be expected to affect oxidative damage, as these processes are cumulative and can lead to long-term consequences [@rice2002brain; @terman2006oxidative].

#### Oxidative stress brought about by experiencing early life CORT are region- and temperature- dependent
We found that DNA damage and lipid peroxidation were affected in region-specific ways. Early conditions did not significantly influence oxidative stress in the olfactory bulbs, but had a pronounced effect on DNA damage and lipid peroxidation in the optic tecta.

CORT exposure reduced DNA damage in optic tecta from cold-incubated lizards but had no effect at higher incubation temperatures. This pattern suggests that CORT might play a protective role under specific thermal conditions. Since CORT helps mobilize energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000], it is possible that under colder incubation conditions — where metabolic rates are expected to be lower — energy mobilised by CORT may be allocated towards antioxidant activity or DNA repairing mechanisms [@kawamura1991glucocorticoid; but see @mcintosh1998glucocorticoids]. In contrast, at higher temperatures, the increased metabolic demands may limit the availability of resources for such protective mechanisms [see @kim2019carry].

In comparison, increased prenatal CORT resulted in higher lipid peroxidation in cold incubated lizards but reduced it in those incubated at warm temperatures. Although the two hormonal treatments did not significantly differ within each incubation temperature, the direction of the CORT effect was the opposite between temperatures. These contrasting CORT–temperature effects on DNA damage and lipid peroxidation suggest that the pathways protecting or repairing each molecule may vary in their thermal sensitivity. For instance, antioxidant enzymes differ in their response to hibernation in the soft-shelled turtle (_Pelodiscus sinensis_), indicating that thermal sensitivity varies among antioxidant system components [@tang2021antioxidant]. An alternative mechanism might involve temperature-dependent changes in membrane lipid composition or fluidity, which may affect how lipids respond to oxidative stress [@zeis2019temperature]. For example, lipid peroxidation levels increased in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids [@zeis2019temperature]. Therefore, temperature-induced structural differences may influence the vulnerability of different molecules to oxidative stress.

@@

### Declaration of AI use
We declare Chat GPT was used mainly for questions related to coding and data analyses. All other parts of the manuscript were written by the authors and those parts where AI was used were checked and modified accordingly by the authors.

>>>>

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. However, early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects. Here, we investigated the combined influence of prenatal corticosterone (CORT)-a key stress hormone in reptiles-and incubation temperature on mitochondrial function and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta-two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta but not the olfactory bulbs. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms linking early-life stress, cell metabolism, and cognitive performance.

@@

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental inputs during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Long-lasting effects on energy production and oxidative stress in the brain may be a key mechanism by which cognitive function is impacted by developmental environments [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks correlate negatively with the number of round-shaped mitochondria-associated with higher oxidative stress-per presynaptic terminal. In conclusion, findings from different studies suggest that mitocondrial activity can be essential in shaping cognitive abilities. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

@@

## Discussion
We examined the impact of prenatal temperature and CORT on visual and chemical perception and determined whether metabolic function underpinned any perceptual differences, as previously shown in mammals [@hara_presynaptic_2014; @chaudhari2022early]. Overall, metabolic processes appeared largely resilient to prenatal environment. However, early-life conditions produced region- and context-dependent effects on oxidative damage and prey detection. We observed interactive effects of prenatal CORT and incubation temperature on lipid proxidation in both regions, while the interaction affected DNA damage only in the optic tecta. Behaviourally, CORT enhanced detection of chemical-but not visual-prey stimuli. Yet, we found no evidence that metabolic explained behavioural variation. Below, we discuss these findings in the context of mitochondrial metabolic function, oxidative stress, and perception.

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. These results were consistent when the effect on the average mitochondrion was analysed (i.e. when metabolic capacity and ROS production were divided by mitochondrial density). Our findings suggest that metabolic function in _L. delicata_'s brain is relatively resilient to early-life conditions. However, evidence from other species indicates that the effects of CORT and temperature can be region-specific: incubation temperature alters metabolic capacity in some, but not all, brain nuclei of leopard geckos (_Eublepharis macularius_) [@coomber_independent_1997], and prenatal stress reduces mitochondrial efficiency in select nuclei of rats (_Rattus norvegicus_) [@gong_chronic_2011]. Thus, while our results point to robustness in the regions studied, other brain areas in _L. delicata_ may be more sensitive to prenatal CORT and temperature.

Alternatively, the effects of early environmental conditions on metabolic function and ROS production may be immediate rather than long-lasting. Mitochondria are highly responsive to environmental changes and energetic demands [@gyllenhammer2020developmental], and any alterations present during incubation may have dissipated once all animals developed under the same post-hatching environment. Nonetheless, even temporary changes in metabolic function during earlier stages would still be expected to affect oxidative damage, as these processes are cumulative and can lead to long-term consequences [@rice2002brain; @terman2006oxidative].

#### Oxidative stress brought about by experiencing early life CORT are region- and temperature- dependent
We found that DNA damage and lipid peroxidation were affected in region-specific ways. Early conditions did not significantly influence DNA damage in the olfactory bulbs, but they did alter lipid peroxidation in both brain nuclei studied. In the optic tecta, prenatal CORT reduced DNA damage in cold-incubated lizards but had no effect at higher incubation temperatures. In both regions, CORT also interacted with incubation temperature to influence lipid peroxidation, increasing it in CORT-treated cold-incubated lizards compared to the CORT-Hot group.

These contrasting patterns suggest that prenatal CORT and temperature may shape oxidative damage through molecule-specific protective pathways, potentially by upregulating different antioxidant systems. Glucocorticoids and temperature each appear to influence antioxidant systems in enzyme-specific ways rather than uniformly across pathways. For example, prenatal stress in rats increases Cu/Zn superoxide dismutase (SOD) and glutathione peroxidase (GPx) activity in the brain while catalase (CAT) was unaffected [@mcintosh1998glucocorticoids]. Similarly, in the soft-shelled turtle (_Pelodiscus sinensis_), CAT expression rises during arousal compared to hibernation, whereas GPx activity remains stable across these temperature-associated states [@tang2021antioxidant]. Such differential regulation could explain why CORT reduces DNA damage under cold incubation but reduces lipid peroxidation under warm incubation, if distinct antioxidant pathways are preferentially activated under different environmental conditions.

Alternative non-exclusive mechanisms could also contribute. CORT mobilises energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000]. Under cold incubation-when metabolic rates are expected to be lower—energy may be allocated towards antioxidant activity or DNA repair [@kawamura1991glucocorticoid; but see @mcintosh1998glucocorticoids]. In contrast, at warmer temperatures, increased metabolic demands may limit the resources available for such protective processes [see @kim2019carry]. These context-dependent changes may explain lower DNA damage in the optic tecta of CORT-Cold lizards, but they do not fully account for the reduced lipid peroxidation in CORT-Hot animals. Additional factors, such as temperature-dependent differences in membrane lipid composition [@zeis2019temperature], may further modulate molecule-specific sensitivity to oxidative stress. For example, lipid peroxidation levels increased in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids, which are more susceptible to oxidative damage [@zeis2019temperature]. Future studies should examine the effects of prenatal CORT and temperature on metabolic rate, antioxidant regulation, and molecule structure and composition in differen brain nuclei of reptiles.

@@

### Declaration of AI use
We declare Chat GPT was used for checking coding errors. All other parts of the manuscript were written by the authors and those parts where AI was used were checked and modified accordingly by the authors.
